Functions and Relations

### 2023-10-21 rec08/paper.tex
cantor thm

Intuitively, a function from a set $A$ to another set $B$ is a rule for assigning each element of $A$ a unique element of $B$.

@@

>>>>

Intuitively, a function is a rule for assigning each element of a set to a unique element of another set.
In set theory, functions can be encoded as a special binary relation.

\begin{defn}
  A binary relation $R \subseteq A \times B$ is (left) \emph{total} if
  \[
    \forall a \in A.\exists b \in B.\,(a,b) \in R
  \]
\end{defn}

\begin{defn}
  A binary relation $R \subseteq A \times B$ is \emph{functional} if
  \[
    \forall a \in A.\forall b \in B.\forall c \in C.\,((a,b) \in R \wedge (a,c) \in R \imp b = c)
  \]
\end{defn}

\begin{defn}
  A function $f$ from a set $A$ to another set $B$, denoted $f : A \to B$ is a binary relation
  \[
    f \subseteq A \times B
  \]
  such that $f$ is \emph{total} and \emph{functional}.
  We write $f(a) = b$ for $(a,b) \in f$.
  Writing the two conditions in this notation is perhaps more illuminating:
  \begin{enumerate}
  \item Totality:
    \[
      \forall a \in A.\exists b \in B.\,f(a) = b
    \]
  \item Functionality:
    \[
      \forall a \in A.\forall b \in B.\forall c \in C.\,(f(a) = b \wedge f(a) = c \imp b = c)
    \]
  \end{enumerate}
  The set $A$ is called the \emph{domain} of $f$, and the set $B$ is called the \emph{codomain} of $f$.
\end{defn}

\begin{thm}[Functional Extensionality]
  Two functions $f,g : A \to B$ are equal if and only if $f(a) = g(a)$ for all $a \in A$.
\end{thm}
\begin{proof}
  The ``only if'' direction is obvious.
  For the ``if'' direction, assume that $f(a) = g(a)$ for all $a \in A$.
  To prove that $f = g$, it suffices to prove $f \subseteq g$ and $g \subseteq f$.
  Now, suppose that $(a,b) \in f$.
  Since $f(a) = g(a)$, $(a, g(a)) \in f$.
  By functionality, $g(a) = b$.
  Thus, $(a,b) \in g$, proving that $f \subseteq g$.
  The proof of $g \subseteq f$ is completely analogous.
\end{proof}

\begin{defn}
  Given two functions $f : A \to B$ and $g : B \to C$, the composition $g \circ f : A \to C$ (reads ``$g$ after $f$'') is a function defined by
  \[
    (g \circ f)(x) = g(f(x))
  \]
  Note that $g \circ f$ is defined only if the codomain of $f$ and the domain of $g$ are the same.
\end{defn}

\begin{lem}
  Composition is associative, i.e., $(f \circ g) \circ h = f \circ (g \circ h)$.
\end{lem}
\begin{proof}
  Exercise.
  Hint: Use functional extensionality.
\end{proof}

\begin{defn}
  For any set $S$, there is a special function $\id_{S}$, called the \emph{identity function on $S$}, defined by
  \[
    \id_{S}(s) = s
  \]
\end{defn}

\begin{lem}
  For any function $f : A \to B$, $\id_{B} \circ f = f$ and $f \circ \id_{A} = f$.
\end{lem}
\begin{proof}
  Exercise.
\end{proof}

\begin{defn}
  A function $f : A \to B$ is \emph{injective}, denoted $f : A \mono B$, if
  \[
    \forall a \in A.\forall a' \in A.\,(f(a) = f(a') \imp a = a')
  \]
\end{defn}

\begin{defn}
  A function $f : A \to B$ is \emph{surjective}, denoted $f : A \epi B$, if
  \[
    \forall b \in B.\exists a \in A.\,f(a) = b
  \]
\end{defn}

\begin{thm}[Cantor's Theorem]
  For any set $S$, there is no surjective functions $f : S \epi \cP(S)$.
\end{thm}
\begin{proof}
  Suppose that $f : S \epi \cP(S)$.
  Consider the subset $\{s \in S \mid s \notin f(s)\}$.
  Since $f$ is surjective, there must be some $s' \in S$ so that $f(s') = \{s \in S \mid s \notin f(s)\}$.
  If $s' \in f(s')$, then by definition, $s' \notin f(s')$.
  Similarly, if $s' \notin f(s')$, then by definition, $s' \in f(s')$.
  This is a contradiction.
\end{proof}

\begin{defn}
  A function $f : A \to B$ is \emph{bijective} if it is injective and surjective.
\end{defn}

@@

\begin{defn}
  A set $S$ is \emph{countable} if there is a bijection $f : S \to \dN$.
\end{defn}
